---
title: "Run Association models"
output: html_notebook
---

```{r}
library(tidyverse)
library(data.table)
library(rstatix)
library(parallel)
library(doParallel)
library(ape)
```

```{r}
# Load covariates
rootDir <- "/workdir/coh22/poaceae_tfbs/"
metadata <- read.table(paste0(rootDir, "data/Poaceae_metadata_filtered_2025.04.16.tsv"), sep = "\t", header = T)
env <- read.table(paste0(rootDir, "output/modelData/envData_759Poaceae_20240821.txt")) %>%
  mutate(assemblyID = rownames(.)) %>%
  dplyr::filter(assemblyID %in% metadata$assemblyID)
EnvTraitMatrix <- env[,286:295] # first 10 ePCs
rownames(EnvTraitMatrix) <- env$assemblyID

phyloK <- read.table(paste0(rootDir, "output/modelData/phyloK_728Poaceae_astral_20250407.txt"))
tree <- read.tree(paste0(rootDir, "output/modelData/PoaceaeTree_angiosperm353_astral_filtered_20250407.nwk"))
colnames(phyloK) <- rownames(phyloK)
```

Calculate environmental residuals for ePC1 after controlling for phylogeny
``` {r}
library(asreml)

# Write resids to file
envPC1_resids <- residuals(envPC1_phyloK)
names(envPC1_resids) <- env$assemblyID
write.table(envPC1_resids, file = "/workdir/coh22/andro_tfbs/results/models/envPC1_resids.txt", col.names = F, quote = F)
```

``` {r}
# Load motif counts
source(paste0(rootDir, "src/07_motifTurnover/load_motif_data.R"))  # Adjust the filename as needed
mat.list <- load_motif_data(
  unenriched_clusters_path = paste0(rootDir, "lists/unenriched_clusters.txt"),
  cluster_meta_path = paste0(rootDir, "data/JASPAR2024_SummarizedClusters.txt"),
  motifs_dir = paste0(rootDir, "output/motifOutput/motifs_by_orthogroup/summarized_by_OG_500")
  #motifs_dir = paste0(rootDir, "output/motifOutput/motifs_by_orthogroup/summarized_by_OG_1000")
)

filtered_list <- mat.list[sapply(mat.list, function(df) nrow(df) >= 200)] # Retain only OGs with at least 200 taxa represented to avoid singularities
```

Run OG-specific association models with asreml
```{r}
Run_MLMs_OrthoSpecific <- function(Kphylo, EnvTraitMatrix, motifMatList) {
  # prepare log
  logfile <- "/workdir/coh22/poaceae_tfbs/output/output_log.txt"
  if (file.exists(logfile)) file.remove(logfile)
  log_to_file <- function(msg) {
    ts <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
    write(paste(ts, msg), file = logfile, append = TRUE)
  }

  # load libs once
  library(dplyr); library(data.table); library(asreml)
  asreml.options(verbose = FALSE)

  ePC_names <- colnames(EnvTraitMatrix)
  n_traits  <- length(ePC_names)
  nOrtho    <- length(motifMatList)
  nCores    <- max(parallel::detectCores() - 1, 1)

  # parallel over OGs
  results_list <- parallel::mclapply(seq_len(nOrtho), function(i) {
    ogID <- names(motifMatList)[i]
    log_to_file(paste("[OG]", ogID, "– starting"))

    motifMatrix <- motifMatList[[i]]
    if (is.null(motifMatrix)) {
      return(list(
        list(ogID, NA_character_, NA_real_,
             rep(NA_real_, n_traits),
             rep(NA_real_, n_traits))
      ))
    }

    # subset to shared taxa
    common_rows <- sort(Reduce(base::intersect, list(
      rownames(motifMatrix),
      rownames(Kphylo),
      rownames(EnvTraitMatrix)
    )))
    Env_sub   <- EnvTraitMatrix[common_rows, , drop = FALSE]
    motif_sub <- motifMatrix[common_rows, , drop = FALSE]
    K_sub     <- Kphylo[common_rows, common_rows, drop = FALSE]

    # build data frame
    input_df <- cbind(as.data.frame(Env_sub), as.data.frame(motif_sub))
    input_df$assemblyIDs <- factor(common_rows)
    data.table::setnames(input_df, gsub("-", "_", names(input_df)))
    colnames(input_df)[seq_len(n_traits)] <- ePC_names

    # per‐motif models
    motifModels <- lapply(seq_len(ncol(motif_sub)), function(j) {
      motifID <- colnames(input_df)[n_traits + j]
      if (length(unique(motif_sub[, j])) <= 1) {
        return(list(ogID, motifID, NA_real_,
                    rep(NA_real_, n_traits),
                    rep(NA_real_, n_traits)))
      }
      tryCatch({
        form   <- as.formula(paste(motifID, "~", paste(ePC_names, collapse = " + ")))
        model  <- asreml(fixed = form,
                         random = ~ vm(assemblyIDs, K_sub),
                         ai.sing = TRUE,
                         data = input_df)
        summ   <- summary(model)
        wald   <- wald.asreml(model)
        phylo  <- summ$varcomp$component[1]
        resid  <- summ$varcomp$component[2]
        frac   <- phylo / (phylo + resid)
        coeffs <- model$coefficients$fixed[2:(1 + n_traits)]
        pvals  <- -(pchisq(wald[2:(1 + n_traits), 3], df = 1,
                           log.p = TRUE, lower.tail = FALSE) / log(10))
        coeffs <- if (length(coeffs) == n_traits) coeffs else rep(NA_real_, n_traits)
        pvals  <- if (length(pvals)  == n_traits) pvals  else rep(NA_real_, n_traits)
        list(ogID, motifID, frac, coeffs, pvals)
      }, error = function(e) {
        log_to_file(paste("[Error] OG=", ogID,
                          "motif=", motifID, "–", e$message))
        list(ogID, motifID, NA_real_,
             rep(NA_real_, n_traits),
             rep(NA_real_, n_traits))
      })
    })

    if (length(motifModels) == 0) {
      motifModels <- list(
        list(ogID, NA_character_, NA_real_,
             rep(NA_real_, n_traits),
             rep(NA_real_, n_traits))
      )
    }
    motifModels
  }, mc.cores = nCores)

  # flatten into one list
  do.call(c, results_list)
}

models_phyloK_envPC1.10_poaceae <-
  Run_MLMs_OrthoSpecific(phyloK, EnvTraitMatrix, filtered_list)

save(models_phyloK_envPC1.10_poaceae_1000, file="/workdir/coh22/poaceae_tfbs/output/models/models_phyloK_envPC1.10_poaceae_1000bpUpstream.RData")
```

```{r}
# Load models
n_pc <- 10

load("/workdir/coh22/poaceae_tfbs/output/models/models_phyloK_envPC1.10_poaceae_500bpUpstream.RData")

cat_models <- lapply(models_phyloK_envPC1.10_poaceae, unlist)

model_output <- mclapply(cat_models, function(x) {
  x <- t(x)
  x <- as.data.frame(x)
  names(x) <- c("OG", "motif", "phylo_explained", paste0("coef_envPC_", 1:n_pc), paste0("log10pval_envPC_", 1:n_pc))
  return(x)
}, mc.cores = detectCores() - 1)

# Identify models in which fewer than ten taxa contain the motif (to filter out from results)
sparse_motifs <- do.call(rbind, lapply(names(filtered_list), function(matrix_name) {
  mat     <- filtered_list[[matrix_name]]
  n_taxa  <- nrow(mat)
  # number of taxa that have any non-zero value for each motif
  presence_counts <- colSums(mat > 0)
  # need at least 10 taxa with the motif, so anything seen in fewer than this is “sparse”
  min_obs <- 10
  low_cols <- which(presence_counts < min_obs)
  
  if (length(low_cols) > 0) {
    data.frame(
      OG     = matrix_name,
      motif  = colnames(mat)[low_cols],
      stringsAsFactors = FALSE
    )
  } else {
    NULL
  }
}))

results <- rbindlist(model_output, fill = T)# %>%
head(results)
results_df <- results %>%
  pivot_longer(
    cols = starts_with("coef") | starts_with("log10pval"),
    names_to = c(".value", "envPC"),
    names_pattern = "(coef|log10pval)_envPC_(\\d+)"
  ) %>%
  arrange(OG, motif, envPC) %>%
  as.data.frame() %>%
  mutate(motif = recode(motif,"SMZ_RAP2_7_TOE2" = "SMZ_RAP2-7_TOE2")) %>%
  dplyr::anti_join(sparse_motifs, by = c("OG", "motif"))  %>% # filter out sparse motif-OG combinations
  na.exclude()


results_df[,c(3,5,6)] <- lapply(results_df[,c(3,5,6)], as.numeric)
results_df$pval <- 10^(-results_df$log10pval)
results_df$p.fdr <- p.adjust(results_df$pval, method = "fdr")
sorted_results <- arrange(results_df, desc(log10pval))


# How many motifs/OGs are significant?
sig_motif_ogs <- dplyr::filter(results_df, p.fdr < 0.01) %>%
  dplyr::select(c(OG, motif)) %>%
  unique()

# Compare p value boxplots across envPCs
ggplot(results_df, aes(x = factor(envPC), y = log10pval)) +
  geom_boxplot() +
  labs(
    title = "Distribution of -log10(p) values by envPC",
    x = "envPC",
    y = "-log10(p)"
  ) +
  theme_minimal()

percentiles <- cut(results_df$pval, breaks = seq(0, 1, by = 0.001), include.lowest = TRUE, right = FALSE)

# Tabulate and plot
barplot(table(percentiles), main = "P-Value Distribution (OG-specific motif occurrence)", xlab = "Per mille", ylab = "Count")

chisq_vals <- qchisq(1 - results_df$pval, df = 1)

# Calculate lambda
lambda_gc <- median(chisq_vals) / qchisq(0.5, df = 1)
```

Run permulations on 1. the top hits 2. all models, comparing empirical vs permulated values
``` {r}
nCores <- max(detectCores() - 1, 1)

top_model_list <- dplyr::filter(results_df, p.fdr < 0.01) %>%
  dplyr::select(c("OG", "motif", "envPC")) %>%
  split(seq(nrow(.))) # produces a list of rows

set.seed(0)
random.100.model.list <- slice_sample(results_df, n = 100, replace = F) %>%
  dplyr::select(c("OG", "motif", "envPC")) %>%
  split(seq(nrow(.))) # produces a list of rows

Run_Permulations <- function(top_model_list, EnvTraitMatrix, motifMatList, tree, n_perm, model_threads = 1, perm_threads = 1) {
  source(file.path(rootDir, "src/09_associationModeling/Permulate.R"))
  library(parallel); library(geiger); library(ape)
  library(dplyr);   library(data.table); library(asreml)
  asreml.options(verbose = FALSE)

  logfile <- file.path(rootDir, "output/perm_output_log.txt")
  if (file.exists(logfile)) file.remove(logfile)
  log_to_file <- function(msg) {
    ts <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
    write(paste(ts, msg), logfile, append = TRUE)
  }


  # Parallelize across motif-og combinations
  results <- mclapply(seq_along(top_model_list), function(i) {
    mdl             <- top_model_list[[i]]
    ogID            <- mdl[["OG"]]
    focal_motif     <- mdl[["motif"]]
    motif_clean     <- gsub("-", ".", focal_motif)
    focal_envPC     <- paste0("envPC_", mdl[["envPC"]])
    log_to_file(paste("Running permulations for", ogID, focal_motif, focal_envPC))

    # retain the matching taxa across datasets
    motifVec   <- motifMatList[[ogID]][, focal_motif]
    common_rows<- intersect(names(motifVec), rownames(EnvTraitMatrix)) %>% sort()

    Env_sub    <- EnvTraitMatrix[common_rows, , drop = FALSE]
    motif_sub  <- motifVec[common_rows]
    K_sub      <- phyloK[common_rows, common_rows, drop = FALSE]
    tree_sub   <- drop.tip(tree, setdiff(tree$tip.label, common_rows))
    tree_rt    <- midpoint.root(tree_sub)
    tree_rt$edge.length[is.na(tree_rt$edge.length)] <- 0.1
    tree_res   <- multi2di(tree_rt)

    # build input_df
    input_df <- data.frame(
      OG         = ogID,
      assemblyID = factor(common_rows),
      motif      = motif_sub,
      Env_sub,
      stringsAsFactors = FALSE
    )
    colnames(input_df)[3] <- motif_clean

    # run permulate with error‐logging
    tryCatch(
        permulate(input_df, motif_clean, focal_envPC, tree_res, K_sub,
            n_perm, perm_threads), # perm_threads = how many permutations to run in parallel
      error = function(e) { log_to_file(e$message); NULL }
    )
  }, mc.cores = model_threads) # How many models to run in parallel

  # name by OG and return
  names(results) <- vapply(top_model_list, `[[`, "", "OG")
  results
}

# Run 1000 permulations of each top hit for significance testing
perm_top_hits <- Run_Permulations(top_model_list, 
                                  EnvTraitMatrix, 
                                  filtered_list, 
                                  tree, 
                                  n_perm = 1000, 
                                  model_threads = 5, 
                                  perm_threads = 20)
#save(perm_top_hits, file = paste0(rootDir, "output/models/permulated_1k_topHits_500bpUpstream.RData"))


 # Permulate 1000 random models to compare empirical vs permulated distributions
random.100.models <- Run_Permulations(random.100.model.list, 
                                  EnvTraitMatrix, 
                                  filtered_list, 
                                  tree, 
                                  n_perm = 1000, 
                                  model_threads = 5, 
                                  perm_threads = 20)
save(random.100.models, file = paste0(rootDir, "output/models/random.100.models.1kPerm_500bpUpstream.RData"))

```

Look at permulation results
``` {r}
load(paste0(rootDir, "output/models/permulated_1k_topHits_500bpUpstream.RData"))
names(perm_top_hits) <- unlist(lapply(perm_top_hits, function(x) paste0(x[1,"OG"], "_", x[1,"responseVar"], "_", x[1,"focal_predVar"])))
perm_df <- mclapply(perm_top_hits, as.data.frame, mc.cores = detectCores() - 1) %>%
  rbindlist() %>%
  rowwise() %>%
  mutate(across(
      phylo_explained:last_col(),
      ~ as.numeric(.))) %>%
  mutate(
    waldStat = get(paste0("waldStat_", focal_predVar)),
    coef = get(paste0("coef_", focal_predVar)),
    pval = get(paste0("pval_log10_", focal_predVar))
  ) %>%
  ungroup() %>%
  dplyr::select(OG, responseVar, responseVar, focal_predVar, model, phylo_explained, waldStat, coef, pval) %>%
  mutate(model_class = if_else(model == "empirical", "empirical", "perm")) 

perm_pvals <- perm_df %>%
  group_by(OG, responseVar, focal_predVar) %>%
  dplyr::reframe(
    empirical_waldStat = waldStat[model_class == "empirical"],
    pval_permulatedNull = mean(abs(waldStat[model_class == "perm"]) >= abs(empirical_waldStat), na.rm = T)
  ) %>%
  mutate(log10pval_perm = -log10(pval_permulatedNull),
         envPC = str_split_fixed(focal_predVar, "_", 2)[,2]) %>%
  dplyr::select(OG, motif = responseVar, envPC, log10pval_perm) %>%
  left_join(results_df, by = join_by(OG, motif, envPC))

hist(perm_pvals$log10pval_perm, xlim = c(0,3))
abline(v = 2)

## 2. Compare pvalues from 100 random models between original and permulated
load(paste0(rootDir, "output/models/random.100.models.1kPerm_500bpUpstream.RData"))

names(random.100.models) <- unlist(lapply(random.100.models, function(x) paste0(x[1,"OG"], "_", x[1,"responseVar"], "_", x[1,"focal_predVar"])))
perm_df <- mclapply(random.100.models, as.data.frame, mc.cores = detectCores() - 1) %>%
  rbindlist() %>%
  rowwise() %>%
  mutate(across(
      phylo_explained:last_col(),
      ~ as.numeric(.))) %>%
  mutate(
    waldStat = get(paste0("waldStat_", focal_predVar)),
    coef = get(paste0("coef_", focal_predVar)),
    pval = get(paste0("pval_log10_", focal_predVar))
  ) %>%
  ungroup() %>%
  dplyr::select(OG, responseVar, responseVar, focal_predVar, model, phylo_explained, waldStat, coef, pval) %>%
  mutate(model_class = if_else(model == "empirical", "empirical", "perm")) 

perm_pvals <- perm_df %>%
  group_by(OG, responseVar, focal_predVar) %>%
  dplyr::reframe(
    empirical_waldStat = waldStat[model_class == "empirical"],
    pval_permulatedNull = mean(abs(waldStat[model_class == "perm"]) >= abs(empirical_waldStat), na.rm = T)
  ) %>%
  mutate(pval_permulatedNull = ifelse(pval_permulatedNull == 0, 0.001, pval_permulatedNull),
    log10pval_perm = -log10(pval_permulatedNull),
    envPC = str_split_fixed(focal_predVar, "_", 2)[,2]) %>%
  dplyr::select(OG, motif = responseVar, envPC, log10pval_perm) %>%
  left_join(results_df, by = join_by(OG, motif, envPC)) %>%
  na.exclude()

# Coerce any inf values to minimum possible from 1000 permulations (0.001)

ggplot(perm_pvals, aes(x = log10pval, y = log10pval_perm)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    x = "Original p-value (-log10)",
    y = "Permulation-derived p-value (log10)",
    title = "Permulated vs empirical p-values"
  ) +
  theme_minimal() +
  xlim(0,5) +
  ylim(0,5)

# Compute spearman correlation between models 
spearman <- cor.test(
  perm_pvals$log10pval,
  perm_pvals$log10pval_perm,
  method = "spearman",
  use = "complete.obs"
)
# 95% correlation


# 2. Scatterplot with y=x line


```

Compare 500bp vs 1000bp upstream results
``` {r}
load("/workdir/coh22/poaceae_tfbs/output/models/models_phyloK_envPC1.10_poaceae_1000bpUpstream.RData")

cat_models_1000bp <- models_phyloK_envPC1.10_poaceae %>%
   lapply(unlist)

model_output_1000bp <- mclapply(cat_models_1000bp, function(x) {
  x <- t(x)
  x <- as.data.frame(x)
  names(x) <- c("OG", "motif", "phylo_explained", paste0("coef_ePC", 1:n_pc), paste0("log10pval_ePC", 1:n_pc))
  return(x)
}, mc.cores = detectCores() - 1)

# Identify models in which ten or fewer taxa contain the motif (to filter out from results)
sparse_motifs_1000bp <-  do.call(rbind, lapply(names(filtered_list), function(matrix_name) {
  mat     <- filtered_list[[matrix_name]]
  n_taxa  <- nrow(mat)
  # number of taxa that have any non-zero value for each motif
  presence_counts <- colSums(mat > 0)
  # need at least 10 taxa with the motif, so anything seen in fewer than this is “sparse”
  min_obs <- 10
  low_cols <- which(presence_counts < min_obs)
  
  if (length(low_cols) > 0) {
    data.frame(
      OG     = matrix_name,
      motif  = colnames(mat)[low_cols],
      stringsAsFactors = FALSE
    )
  } else {
    NULL
  }
}))

results_1000bp <- rbindlist(model_output_1000bp, fill = T)# %>%
head(results_1000bp)
results_df_1000bp <- results_1000bp %>%
  pivot_longer(
    cols = starts_with("coef") | starts_with("log10pval"),
    names_to = c(".value", "ePC"),
    names_pattern = "(coef|log10pval)_ePC(\\d+)"
  ) %>%
  arrange(OG, motif, ePC) %>%
  as.data.frame() %>%
  dplyr::filter(OG %in% names(filtered_list)) %>% # Only retain OGs with enough taxa
  mutate(motif = recode(motif,"SMZ_RAP2_7_TOE2" = "SMZ_RAP2-7_TOE2")) %>%
  dplyr::anti_join(sparse_motifs, by = c("OG", "motif"))  # filter out sparse motif-OG combinations


results_df_1000bp[,c(3,5,6)] <- lapply(results_df_1000bp[,c(3,5,6)], as.numeric)
results_df_1000bp$pval <- 10^(-results_df_1000bp$log10pval)
results_df_1000bp$p.fdr <- p.adjust(results_df_1000bp$pval, method = "fdr")
sorted_results_1000bp <- arrange(results_df_1000bp, desc(log10pval))

comparison.df <- dplyr::select(sorted_results_1000bp, c(OG, motif, ePC, log10pval_1000bp = log10pval)) %>%
  inner_join(results_df) %>%
  dplyr::rename(log10pval_500bp = log10pval)

ggplot(comparison.df, aes(x = log10pval_500bp, y = log10pval_1000bp)) +
  stat_bin2d(bins = 200, aes(fill = ..count..)) +
  scale_fill_viridis_c(
    option = "inferno",
    trans  = "sqrt"
  ) +
  geom_smooth(method = "lm", color = "steelblue", se = FALSE, size = 1) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid       = element_blank(),
    legend.position  = "none"
  ) +
  labs(
    x     = expression(log[10]~pval[500~bp]),
    y     = expression(log[10]~pval[1000~bp]),
    title = "Motif x Orthogroup association model comparison") + 
     xlim(-1,50) +
     ylim(-1,50)

# Compute spearman correlation between models
spearman <- cor.test(
  comparison.df$log10pval_500bp,
  comparison.df$log10pval_1000bp,
  method = "spearman",
  use = "complete.obs"
)

sig_motif_ogs_1kb <- dplyr::filter(results_df_1000bp, p.fdr < 0.01) %>%
  dplyr::select(c(OG, motif)) %>%
  unique()

# How many in common btwn models
top_hits_intersection <- inner_join(sig_motif_ogs, sig_motif_ogs_1kb)
# Compare p value boxplots across envPCs
ggplot(results_df_1000bp, aes(x = factor(ePC), y = log10pval)) +
  geom_boxplot() +
  labs(
    title = "Distribution of -log10(p) values by ePC",
    x = "ePC",
    y = "-log10(p)"
  ) +
  theme_minimal()
```
Function to make QQ plot 
```{r}
PlotQQ <- function(df) {
  # Create observed vs expected p values for plot
  observed_pvals = sort(df$log10pval, decreasing = T)
  observed_p.fdr = p.adjust(10^(-observed_pvals), method = "fdr")
  expected_pvals = qunif(seq(0, 1, length.out = length(observed_pvals) + 2), min = 0, max = 1)  ## Generate expected values. Note that we are using length+2
  expected_pvals = expected_pvals[expected_pvals != 0 & expected_pvals != 1]  ## Remove the two extra values since they are 0 and 1

  p_df = data.frame(observed = observed_pvals,
                    expected = -log10(expected_pvals),
                    color = ifelse(observed_p.fdr < 0.01, "goldenrod1", "#2C7BB6"))  # Assign color based on p.fdr
  
  ggplot(p_df, aes(x = expected, y = observed, color = color)) +
    geom_point(size = 7, alpha = 0.8) +  # Color determined by 'color' column
    scale_color_identity() +  # Use color values directly
    geom_abline(intercept = 0, slope = 1, color = '#D7191C', linetype = 'dashed', linewidth = 0.8) +  # Dashed red line
    labs(
      x = expression("-log"[10]~"Expected p-value"),
      y = expression("-log"[10]~"Observed p-value")
    ) +
    theme_minimal(base_size = 36) +  # Minimal theme for clean appearance
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),  # Center and bold title
      axis.title = element_text(face = "bold"),  # Bold axis titles
      panel.grid.major = element_line(linewidth = 0.5, color = 'grey80'),  # Light gray grid lines
      panel.grid.minor = element_blank()  # Remove minor grid lines
    )
}

PlotQQ_sampled <- function(results_df, frac = 0.01) {
  df_sampled <- results_df[sample(nrow(results_df), size = max(10000, round(nrow(results_df) * frac))), ]  # Keep max(10000 or frac)
  PlotQQ(df_sampled)
}

PlotQQ_sampled(results_df)

# Save full
ggsave("/workdir/coh22/poaceae_tfbs/results/qqplot_envPC1_10_1kbupstream_perOG_p01.png", plot = PlotQQ(results_df_1000bp), width = 10, height = 10, dpi = 300)
```

Look at top hits
``` {r}
maize_gene_desc <- read.table("/workdir/coh22/poaceae_tfbs/output/go_enrichment/B73_DE.out", sep = "\t", header = TRUE, quote="", fill=TRUE) %>%
  dplyr::select(c(qpid, desc)) %>%
  mutate(ZmID = sapply(qpid, function(x) strsplit(x, "_")[[1]][1])) %>%
  group_by(ZmID) %>%
  dplyr::summarise(desc = paste(unique(desc), collapse = "; ")) %>%
  mutate(desc = str_replace(desc, "\t", ";"))

OGkey <- read.table("/workdir/coh22/poaceae_tfbs/output/ogs/OGToZm_mapping_v2.txt") %>%
  dplyr::rename(c(ZmID = V1, OG = V2)) %>%
  mutate(ZmID = sapply(ZmID, function(x) strsplit(x, "_")[[1]][1])) %>%
  left_join(maize_gene_desc)

View(left_join(sorted_results[1:1000,], OGkey))

focal_ePC <- dplyr::filter(results_df, ePC == "1")
#focal_ePC <- dplyr::filter(results_df, ePC == "2")
PlotQQ(focal_ePC)
focal_ePC[order(focal_ePC$log10pval, decreasing = T),][1:100,] %>%
  left_join(OGkey) %>%
  write.table("/workdir/coh22/poaceae_tfbs/results/models/envPC1_top_hits_OGspecific.txt", quote = T, row.names = F, sep = "\t")

# How many have pvals <1e-5?

#top_hits <- dplyr::filter(results_df, log10pval > 5) %>%
bin_length <- round(nrow(results_df) * 0.001) # top 0.1% of hits
top_hits <- results_df_1000bp[order(results_df_1000bp$log10pval, decreasing = T),][1:nrow(sig_motif_ogs_1kb),] %>%
#top_hits <- focal_ePC[order(focal_ePC$log10pval, decreasing = T),][1:100,] %>%
  left_join(OGkey)

View(top_hits)

write.csv(top_hits, "/workdir/coh22/poaceae_tfbs/results/top_hits_OGspecific_1kbUpstream.csv", quote = T, row.names = F)
```

Plot candidate motifs/OGs on phylogeny
``` {r}
source("/workdir/coh22/poaceae_tfbs/src/08_envirotyping/PlotTraitsOnTree.R")
candidate.mat <- as.data.frame(filtered_list["OG0018131"])
env <- read.table("/workdir/coh22/poaceae_tfbs/output/modelData/envData_759Poaceae_20240821.txt") %>%
  mutate(assemblyID = rownames(.))

plotTraitOnTree(trait = candidate.mat[,"OG0018131.HSF_GARP"], taxa = rownames(candidate.mat), tree = tree, visible_tip_labels = F, shape = "phylogram", colors = c("lightgrey", "black"))

candidate.counts <- as.data.frame(cbind(candidate.mat$OG0018131.HSF_GARP, rownames(candidate.mat))) %>%
  setnames(c("motifCount", "assemblyID"))

all_rep_taxa <- left_join(candidate.counts, env)

# For quantitative variation in motif count
ggplot(all_rep_taxa, aes(x = factor(motifCount), y = envPC_1)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5, fill = "lightblue") +
  geom_jitter(width = 0.2, size = 5, alpha = 0.7, color = "blue") +
  geom_smooth(
    aes(x = as.numeric(motifCount), y = envPC_1, group = 1),
    method    = "lm",
    se        = TRUE,
    color     = "seagreen",
    linetype  = "dashed", 
    size      = 1
  ) +
  labs(x = "HSF/GARP Motif Count, OG0018131", y = "Environmental PC1") +
  theme_minimal(base_size = 36)

# Taxa with more than 2 copies of the motif
top_taxa <- filter(all_rep_taxa, as.numeric(motifCount) > 2) %>% 
  left_join(metadata) %>%
  dplyr::select(c(Species, assemblyID, envPC_1, motifCount))
```
``` {r}

all_rep_taxa <- all_rep_taxa %>%
  mutate(motif_group = ifelse(motifCount == 0, "0", "≥1"))

ggplot(all_rep_taxa, aes(x = factor(motif_group, levels = c("0", "≥1")), y = envPC_1)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5, fill = "lightblue") +
  geom_jitter(width = 0.2, size = 5, alpha = 0.7, color = "blue") +
  labs(x = "HSF/GARP Motif Count, OG0018131", y = "Environmental PC1") +
  theme_minimal(base_size = 36)
t_test_result <- t.test(envPC_1 ~ motif_group, data = all_rep_taxa)
t_test_result
```

GO enrichment
```{r}
library(topGO)
mappingFile = read.table("/workdir/coh22/poaceae_tfbs/output/ogs/OGToZm_mapping_v2.txt") %>% # map to maize
   mutate(V1=sapply(V1, function(x) strsplit(x, "_")[[1]][1])) # For maize

all_ogs <- unique(mappingFile[mappingFile[,2]%in%names(mat.list),1])
top_ogs = unique(mappingFile[mappingFile[,2]%in%unique(top_hits$OG),1])

GODB <- readMappings("/workdir/coh22/poaceae_tfbs/output/go_enrichment/Zm00001eb.1.fulldata_goList.txt",IDsep = ",")
background = intersect(names(GODB),all_ogs)
tmp=factor(as.integer(background%in%top_ogs))
names(tmp)=background

# Define a function to process each ontology
algo <- "weight01" # Choose topGO algorithm
process_topGO <- function(ontology, allGenes, GODB, top_ogs, mappingFile) {
  tgd1 <- new("topGOdata", ontology = ontology, allGenes = allGenes, nodeSize = 5, annot = annFUN.gene2GO, gene2GO = GODB)
  resTopGO <- runTest(tgd1, algorithm = algo, statistic = "Fisher")
  GO_res_table <- GenTable(tgd1, pval = resTopGO, orderBy = "pval", 
                           ranksOf = "pval", topNodes = length(resTopGO@score), numChar = 100)
  GO_res_table$sigGenes <- sapply(GO_res_table[, 1], function(x) {
    paste(mappingFile[mappingFile[, 1] %in% intersect(top_ogs, genesInTerm(tgd1, x)[[1]]), 2], collapse = ",")
  })
  GO_res_table_df <- as.data.frame(GO_res_table)
  
  # Filtering and transformation
  significant_results <- GO_res_table_df %>%
    #mutate(Fisher.FDR = p.adjust(as.numeric(pval), method = "fdr")) %>%
    mutate(pval = as.numeric(as.character(pval))) %>%   # make sure it’s numeric
    filter(pval < 0.05) %>%
    mutate(neglog10Fisher = -log10(as.numeric(pval))) %>%
    arrange(desc(neglog10Fisher))
  
  # Add ontology information
  significant_results <- significant_results %>%
    mutate(Ontology = ontology) %>%
    mutate(enrichment = Significant / Expected)
  
  return(significant_results)
}

# Process each ontology
significant_BP <- process_topGO("BP", tmp, GODB, top_ogs, mappingFile)
significant_MF <- process_topGO("MF", tmp, GODB, top_ogs, mappingFile)
significant_CC <- process_topGO("CC", tmp, GODB, top_ogs, mappingFile)

# Merge results
merged_results <- bind_rows(significant_BP, significant_MF, significant_CC) %>%
  arrange(desc(neglog10Fisher))

#View(merged_results_most_conserved[,c(2,4,5,8,9)])
View(merged_results[,c(2,3,4,5,6,8,10)])
top_terms <- merged_results[1:7,]
#top_terms <- significant_MF[1:6,]
top_terms$Term <- sapply(top_terms$Term, function(x) substr(x, 1, 30)) # Truncate terms to make plotting prettier
# bubble plot
ggplot(top_terms, aes(
  x = neglog10Fisher, 
  y = reorder(Term, neglog10Fisher), 
  size = Significant, 
  color = enrichment
)) +
  geom_point(alpha = 0.7) +
  scale_x_continuous(limits = c(1.5, 3)) +
  scale_color_gradient(low = "red", high = "blue", limits = c(1, 12)) +
  scale_size(range = c(10, 40)) +  # This line affects bubble size
  labs(
    x = "-log10(Fisher)", 
    y = NULL, # Remove axis title for y to match your example
    size = "Significant",
    color = "Enrichment"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10), # Adjust text size for better visibility
    axis.title.x = element_text(size = 12),
    legend.position = "right",
    panel.grid.major.y = element_line(size = 0.5, color = "grey80"), # Add gridlines for alignment
    plot.title = element_text(hjust = 0.5, size = 14))

write.csv(merged_results[,c(2,3,4,5,6,8,10)], "/workdir/coh22/poaceae_tfbs/results/models/GO_topAssociationModelHits.csv", quote = F, row.names = F)
```

Associate presence/absence of upstream region in dataset with env phenotypes
``` {r}
# Make empty list to store OG presence/absence
og.presence.absence <- replicate(
  length(mat.list),
  as.data.frame(matrix(NA_real_, nrow = 727, ncol = 3)),
  simplify = FALSE
)
names(og.presence.absence) <- names(mat.list)
# Create presence/absence matrix for  each orthogroup
for(ogIndex in 1:length(mat.list)) {
  og.name <- names(mat.list)[ogIndex]
  # Extract taxa represented in OG
  og.mat <- mat.list[[ogIndex]]
  og.sp.names <- rownames(og.mat)
  
  # Get presence/absence for each motif
  colnames(og.presence.absence[[ogIndex]]) <- c("assemblyID", "OG", "ogPresent")
  rownames(og.presence.absence[[ogIndex]]) <- metadata$assemblyID
  og.presence.absence[[ogIndex]]$assemblyID <- metadata$assemblyID
  og.presence.absence[[ogIndex]]$OG <- og.name
  og.presence.absence[[ogIndex]]$ogPresent <- ifelse(
    og.presence.absence[[ogIndex]]$assemblyID %in% og.sp.names,
    1,
    0
  )
}

# Function to run association models with asreml
AssociateOrthogroupPresence <- function(ogMatList, EnvTraitMatrix) {
  # Load functions and packages once
  library(dplyr)
  library(data.table)
  library(asreml)
  source(paste0(rootDir,"src/09_associationModeling/Permulate.R"))

  asreml.options(verbose = FALSE)

  # Set up logging
  logfile <- paste0(rootDir, "/output/orthogroupPresence_output_log.txt")
  if (file.exists(logfile)) file.remove(logfile)
  log_to_file <- function(msg, logfile) {
    timestamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S")
    write(paste(timestamp, msg), file = logfile, append = TRUE)
  }
  
  cl <- parallel::makeCluster(1)
  doParallel::registerDoParallel(cl)
  
  results <- foreach::foreach(
    i = seq_along(ogMatList),
    .errorhandling = 'pass',
    .packages  = c('dplyr','data.table', 'asreml'),
    .export    = c('phyloK','rootDir','run_model_asreml')
    ) %dopar% {
    ogMat <- ogMatList[[i]]
    ogID <- ogMat$OG[1]
    log_to_file(paste("Running association model for", ogID), logfile)

    ogPresenceVec <- ogMat$ogPresent
    names(ogPresenceVec) <- ogMat$assemblyID
    # prep data for model
    common_rows <- sort(Reduce(base::intersect, list(
      names(ogPresenceVec),
      rownames(EnvTraitMatrix)
    )))

    Env_sub   <- EnvTraitMatrix[common_rows, , drop = FALSE]
    ogPresence_sub <- ogPresenceVec[common_rows]
    K_sub     <- phyloK[common_rows, common_rows, drop = FALSE]
  

    input_df <- data.frame(
      OG = ogID,
      assemblyID = factor(common_rows),
      ogPresence = ogPresence_sub,
      Env_sub
    )

    # Run model with binomial family
    tryCatch({
      run_model_asreml(input_df, K_sub, modelFamily = "binomial")
    }, error = function(e) {
      log_to_file(e$message, logfile)
      NULL
    })
  }

  parallel::stopCluster(cl)
  return(results)
}

# Run association models
og.presence.absence.models <- AssociateOrthogroupPresence(og.presence.absence, EnvTraitMatrix)

save(og.presence.absence.models, file = paste0(rootDir, "output/models/og.presence.absence.models_500bp.Rdata"))

load(file = paste0(rootDir, "output/models/og.presence.absence.models_500bp.Rdata"))
cat_og_models <- lapply(og.presence.absence.models, unlist)

model_output_og_pa <- mclapply(cat_og_models, function(x) {
  x <- t(x)
  x <- as.data.frame(x)
  names(x) <- c("phylo_explained", paste0("waldStat_envPC_", 1:n_pc), paste0("coef_envPC_", 1:n_pc), paste0("log10pval_envPC_", 1:n_pc))
  return(x)
}, mc.cores = detectCores() - 1)

# Identify models in which ten or fewer taxa contain the motif (to filter out from results)
sparse_ogs <- unlist(lapply(names(mat.list), function(matrix_name) {
  mat     <- mat.list[[matrix_name]]
  n_taxa  <- nrow(mat)
  # need at least 10 taxa with the OG, so anything seen in fewer than this is “sparse”
  min_obs <- 10
  low_cols <- which(n_taxa < min_obs)
  
  if (n_taxa < min_obs) {
    matrix_name
  } else {
    NULL
  }
}))

results_og_pa <- rbindlist(model_output_og_pa, fill = T)
results_og_pa$OG <- names(mat.list)
results_df_og_pa <- results_og_pa %>%
  pivot_longer(
    cols = starts_with("waldStat") | starts_with("coef") | starts_with("log10pval"),
    names_to = c(".value", "envPC"),
    names_pattern = "(waldStat|coef|log10pval)_envPC_(\\d+)"
  ) %>%
  arrange(OG, envPC) %>%
  as.data.frame() %>%
  dplyr::filter(!OG %in% names(sparse_ogs)) # Only retain OGs with enough taxa


results_df_og_pa[,c(1,4:6)] <- lapply(results_df_og_pa[,c(1,4:6)], as.numeric)
results_df_og_pa$pval <- 10^(-results_df_og_pa$log10pval)
results_df_og_pa$p.fdr <- p.adjust(results_df_og_pa$pval, method = "fdr")
sorted_results_og_pa <- arrange(results_df_og_pa, desc(log10pval))
PlotQQ(results_df_og_pa)
```



