---
title: "Conservation of motif enrichment within regulatory regions"
author: "Charlie Hale"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(stats)
library(VennDiagram)
library(tidyverse)
library(pheatmap)
library(UpSetR)
```

# Load JASPAR motif metadata
``` {r}
rootDir <- "/workdir/coh22/poaceae_tfbs/"
grass_taxa <- c("Zea mays", "Hordeum vulgare", "Triticum aestivum", "Oryza sativa Japonica Group", "Oryza sativa Indica Group", "Oryza sativa") # to subset motifs that were characterized in grasses
plant_motif_key <- read.table(paste0(rootDir, "data/JASPAR2024_CORE_plants_nr_PFM_key.txt"), header = T, sep = "\t") %>%
  dplyr::rename(motif_id = MatrixID) #%>%
  #dplyr::filter(!SpeciesName %in% grass_taxa)
  # retain only motifs characterized in grasses
plant_motif_list <- dplyr::select(plant_motif_key, motif_id)
```

# Load UMR enrichment results
``` {r}
umr_res_dir <- paste0(rootDir, "output/motif_enrichment/umrs_crisp2020/results/")
brachy <- read.table(paste0(umr_res_dir, "Brachypodium_distachyon_stat_permu_test.txt"), 
                     header = T) %>%
  add_column(species = "brachy") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr")) 

rice <- read.table(paste0(umr_res_dir, "Oryza_sativa_stat_permu_test.txt"),
                   header = T) %>%
  add_column(species = "rice") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))

maize <- read.table(paste0(umr_res_dir, "Zea_mays_stat_permu_test.txt"),
                    header = T) %>%
  add_column(species = "maize") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))

barley <- read.table(paste0(umr_res_dir, "Hordeum_vulgare_stat_permu_test.txt"),
                     header = T) %>%
  add_column(species = "barley") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))

sorghum <- read.table(paste0(umr_res_dir, "Sorghum_bicolor_stat_permu_test.txt"),
                      header = T) %>%
  add_column(species = "sorghum") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))
```



# Plot enrichment results for sorghum
``` {r}
sorghum_unfiltered <- mutate(sorghum, log_p_value = -log10(p_value))
# Coerce infinite values to max
sorghum_unfiltered$log_p_value[is.infinite(sorghum_unfiltered$log_p_value)] <- 350 

ggplot(sorghum_unfiltered, aes(x = fold_change, y = log_p_value)) +
  geom_point(aes(color = p.fdr < 0.01), size = 2) + 
  scale_color_manual(values = c("black", "orange")) +
  theme_minimal() +
  labs(x = "Fold Change", y = "-log10(p-value)", color = "p.fdr < 0.01") +
  ylim(0, 375) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  )
```

``` {r}
df_list <- list(maize, brachy, sorghum, rice, barley)
combined_df <- bind_rows(df_list)

wide_df <- combined_df %>%
  dplyr::select(motif_id, species, fold_change) %>%
  pivot_wider(names_from = species, values_from = fold_change) %>%
  as.data.frame()
rownames(wide_df) <- wide_df$motif_id
wide_df <- dplyr::select(wide_df, -motif_id)

cor_matrix <- cor(wide_df[, -1], use = "pairwise.complete.obs")

pheatmap(cor_matrix, color = colorRampPalette(c("pink", "red"))(100), fontsize = 24)
```
 
# Filter to retain significantly enriched motifs for Brachypodium, sorghum, and rice
``` {r}
filter_sig_enriched <- function(df) {
  dplyr::filter(df, fold_change > 1, p.fdr < 0.01)
}

brachy_enriched <- filter_sig_enriched(brachy)
sorghum_enriched <- filter_sig_enriched(sorghum)
rice_enriched <- filter_sig_enriched(rice)
maize_enriched <- filter_sig_enriched(maize)
barley_enriched <- filter_sig_enriched(barley)

```

# Upset plot of enriched motifs across species
``` {r}
motif_list <- list(
  sorghum = sorghum_enriched$motif_id,
  maize = maize_enriched$motif_id,
  rice = rice_enriched$motif_id,
  brachypodium = brachy_enriched$motif_id,
  barley = barley_enriched$motif_id
)
upset_data <- fromList(motif_list)

upset(upset_data, 
      order.by = "freq", 
      nsets = 5, 
      keep.order = TRUE, 
      text.scale = 2.5, 
      nintersects = 9,
      sets = c("barley", "brachypodium", "rice", "maize", "sorghum"))

upset(upset_data, order.by = "freq", nsets = 4, keep.order = T, text.scale = 2.5, sets = c("barley", "brachypodium", "rice", "maize", "sorghum"))
```

``` {r}
superset_enriched <- unique(c(rice_enriched$motif_id, sorghum_enriched$motif_id, brachy_enriched$motif_id, maize_enriched$motif_id, barley_enriched$motif_id))
intersection <- Reduce(intersect, list(rice_enriched$motif_id, sorghum_enriched$motif_id, brachy_enriched$motif_id, maize_enriched$motif_id, barley_enriched$motif_id))

write.table(intersection, 
            paste0(rootDir,"output/motif_enrichment/umr_enriched_plant_motifs_JASPAR2024.txt"),
            row.names = F, col.names = F, quote = F)

enriched.motifs <- dplyr::filter(plant_motif_key, motif_id %in% intersection)
enriched.clusters <- unique(enriched.motifs$ClusterDesc)
all.clusters <- unique(plant_motif_key$ClusterDesc)
unenriched.clusters <- setdiff(all.clusters, enriched.clusters)

write.table(unenriched.clusters, paste0(rootDir,"lists/unenriched_clusters.txt"),
            quote = F, col.names = F, row.names = F)

```

Look into motifs that are enriched in at least one species, but not all
``` {r}
unshared.enrich.ids <- superset_enriched[!superset_enriched %in% intersection] # get this set
# Which motifs are these?
unshared.enrich <- dplyr::filter(plant_motif_key, motif_id %in% unshared.enrich.ids)
unshared.enrich.names <- unshared.enrich$Name
# See how similar enrichment values are for these across species
unshared.enrich.vals <- wide_df[unshared.enrich.ids,]
rownames(unshared.enrich.vals) <- unshared.enrich.names
log_matrix <- log2(unshared.enrich.vals)

# Plot with fixed log2 scale
png(paste0(rootDir, "results/figs/differentiallyEnrichedMotifs.png"), width = 18, height = 6, units = "in", res = 1500)
heatmap <- pheatmap(t(log_matrix),
         cluster_rows = T,
         cluster_cols = T,
         show_rownames = T,
         scale = "none",
         color = colorRampPalette(c("blue", "white", "red"))(100),
        breaks = seq(-1, 1, length.out = 101),
        fontsize_col = 10,
        fontsize_row = 30)
heatmap
dev.off()
```
# Look into significantly depleted motifs
``` {r}
filter_sig_depleted <- function(df) {
  dplyr::filter(df, fold_change < 1, p.fdr < 0.01)
}

brachy_depleted <- filter_sig_depleted(brachy)
sorghum_depleted <- filter_sig_depleted(sorghum)
rice_depleted <- filter_sig_depleted(rice)
maize_depleted <- filter_sig_depleted(maize)
barley_depleted <- filter_sig_depleted(barley)

depleted.intersection <- Reduce(intersect, list(rice_depleted$motif_id, sorghum_depleted$motif_id, brachy_depleted$motif_id, maize_depleted$motif_id, barley_depleted$motif_id))

depleted.motifs <- dplyr::filter(plant_motif_key, motif_id %in% depleted.intersection) %>%
  left_join(wide_df) %>%
  dplyr::filter(SpeciesName %in% grass_taxa)
```

``` {r}
# Data with ordered factor
df <- data.frame(
  Group = rep(c("Ground truth functional grass motifs", "Non-grass motifs"), each = 3),
  Category = factor(rep(c("Enriched", "Depleted", "Neither/Mixed"), times = 2),
                    levels = c("Neither/Mixed", "Depleted", "Enriched")),
  Percent = c(76, 12, 12, 49.8, 25.3, 24.9)
)

# Plot
ggplot(df, aes(x = Group, y = Percent, fill = Category)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_y_continuous(labels = scales::percent_format(scale = 1), expand = c(0, 0)) +
  labs(x = NULL, y = "Percentage") +
  scale_fill_manual(values = c("Enriched" = "#4E79A7", 
                               "Depleted" = "#E15759", 
                               "Neither/Mixed" = "#BAB0AC")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right",
        panel.grid.major.x = element_blank())
```

## Compare with ACR enrichment
# Load ACR enrichment results
``` {r}
brachy_acr <- read.table("/workdir/coh22/poaceae_tfbs/output/umr_enrichment/atac/Brachypodium_7days_leaf_ACRs.txt", 
                     header = T) %>%
  add_column(species = "brachy") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr")) 

rice_acr <- read.table("/workdir/coh22/poaceae_tfbs/output/umr_enrichment/atac/Rice_7days_leaf_ACRs.txt",
                   header = T) %>%
  add_column(species = "rice") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))

maize_acr <- read.table("/workdir/coh22/poaceae_tfbs/output/umr_enrichment/atac/Maize_7days_leaf_ACRs.txt",
                    header = T) %>%
  add_column(species = "maize") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))

barley_acr <- read.table("/workdir/coh22/poaceae_tfbs/output/umr_enrichment/atac/Barley_7days_leaf_ACRs.txt",
                     header = T) %>%
  add_column(species = "barley") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))

sorghum_acr <- read.table("/workdir/coh22/poaceae_tfbs/output/umr_enrichment/atac/Sorghum_7days_leaf_ACRs.txt",
                      header = T) %>%
  add_column(species = "sorghum") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))

setaria_acr <- read.table("/workdir/coh22/poaceae_tfbs/output/umr_enrichment/atac/Setaria_7days_leaf_ACRs.txt",
                      header = T) %>%
  add_column(species = "setaria") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))

arabidopsis_acr <- read.table("/workdir/coh22/poaceae_tfbs/output/umr_enrichment/atac/Arabidopsis_7days_leaf_ACRs.txt",
                      header = T) %>%
  add_column(species = "arabidopsis") %>%
  filter(motif_id %in% plant_motif_list$motif_id) %>%
  mutate(p.fdr = p.adjust(p_value, method = "fdr"))
```

Correlation between UMR and ACR enrichment
``` {r}
# Merge the data frames by motif_id
merged_df <- merge(maize_acr, maize, by = "motif_id", suffixes = c("_acr", "_maize"))

# Calculate the Pearson correlation between the fold_change columns
correlation_value <- cor(merged_df$fold_change_acr, merged_df$fold_change_maize)

```

``` {r}
# motifs that are only enriched in ACRs
View(dplyr::filter(plant_motif_key, motif_id %in% setdiff(acr_intersection, intersection)))
```

# Filter to retain significantly enriched motifs for Brachypodium, sorghum, and rice
``` {r}
filter_sig_enriched <- function(df) {
  dplyr::filter(df, fold_change > 1, p.fdr < 0.01)
}

brachy_enriched <- filter_sig_enriched(brachy_acr)
sorghum_enriched <- filter_sig_enriched(sorghum_acr)
rice_enriched <- filter_sig_enriched(rice_acr)
maize_enriched <- filter_sig_enriched(maize_acr)
barley_enriched <- filter_sig_enriched(barley_acr)
setaria_enriched <- filter_sig_enriched(setaria_acr)
arabidopsis_enriched <- filter_sig_enriched(arabidopsis_acr)
```

# Upset plot of enriched motifs across species
``` {r}
motif_list <- list(
  sorghum = sorghum_enriched$motif_id,
  maize = maize_enriched$motif_id,
  rice = rice_enriched$motif_id,
  brachypodium = brachy_enriched$motif_id,
  barley = barley_enriched$motif_id,
  arabidopsis = arabidopsis_enriched$motif_id,
  setaria = setaria_enriched$motif_id
)
upset_data <- fromList(motif_list)

upset(upset_data, 
      order.by = "freq", 
      nsets = 5, 
      keep.order = TRUE, 
      text.scale = 2.5, 
      nintersects = 9,
      sets = c("arabidopsis", "barley", "brachypodium", "rice", "setaria","maize", "sorghum"))

```

``` {r}
all_enriched_motifs <- unique(c(rice_enriched$motif_id, sorghum_enriched$motif_id, brachy_enriched$motif_id, maize_enriched$motif_id, barley_enriched$motif_id, arabidopsis_enriched$motif_id, setaria_enriched$motif_id))

all_grass_enriched <- unique(c(rice_enriched$motif_id, sorghum_enriched$motif_id, brachy_enriched$motif_id, maize_enriched$motif_id, barley_enriched$motif_id,setaria_enriched$motif_id))
grass_intersection <- Reduce(intersect, list(rice_enriched$motif_id, sorghum_enriched$motif_id, brachy_enriched$motif_id, maize_enriched$motif_id, barley_enriched$motif_id, setaria_enriched$motif_id))

intersection <- Reduce(intersect, list(rice_enriched$motif_id, sorghum_enriched$motif_id, brachy_enriched$motif_id, maize_enriched$motif_id, barley_enriched$motif_id, arabidopsis_enriched$motif_id, setaria_enriched$motif_id))


# Look at motifs that are grass-specific
grass_specific <- dplyr::filter(plant_motif_key, motif_id %in% grass_intersection[!grass_intersection %in% intersection])
arabidopsis_specific <- dplyr::filter(plant_motif_key, motif_id %in% arabidopsis_enriched$motif_id[!arabidopsis_enriched$motif_id %in% all_grass_enriched])
conserved_motifs <- dplyr::filter(plant_motif_key, motif_id %in% intersection)

# Percentages of motifs in each category
pct_conserved <- length(intersection) / length(all_enriched_motifs) # 57.8%
pct_grass_specific <- length(grass_intersection[!grass_intersection %in% intersection]) / length(all_grass_enriched) # 9.5%
pct_arabidopsis_specific <- length(arabidopsis_enriched$motif_id[!arabidopsis_enriched$motif_id %in% all_grass_enriched]) / length(arabidopsis_enriched$motif_id) # 5.6%
```

``` {r}
#write.table(grass_intersection, 
#            "/workdir/coh22/poaceae_tfbs/output/umr_enrichment/acr_enriched_plant_motifs_JASPAR2024.txt",
#            row.names = F, col.names = F, quote = F)

enriched.motifs <- dplyr::filter(plant_motif_key, motif_id %in% intersection)
enriched.clusters <- unique(enriched.motifs$ClusterDesc)
all.clusters <- unique(plant_motif_key$ClusterDesc) # Fewer clusters are represented in ACRs than in UMRs (31 vs 35)
unenriched.clusters <- setdiff(all.clusters, enriched.clusters) 

#write.table(unenriched.clusters, "/workdir/coh22/poaceae_tfbs/lists/acr_unenriched_clusters.txt",
#            quote = F, col.names = F, row.names = F)
```
